Demes: a specification for interchanging complex demography

### 2022-04-08 paper.tex
changes from kevin's comments

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships (i.e., the topology) or by continuous or discrete migration events.
The deomgraphic model itself omits information about genome biology, such as
genome size and architecture, ploidy, and mutation and recombination rates.
It also omits any information about selection operating within populations.
Within populations, allele-frequency dynamics are assumed to
follow the Wright-Fisher model, so that generations are discrete and
non-overlapping and population sizes are independent of mean fitness
(i.e., ``soft'' selection~\citep{christiansen1975hard}).
While these restrictions preclude some population-genetic scenarios of
interest, such as continuous spatial structure
\aprcomment{citations here, and revisit} or
\aprcomment{stochastic carrying-capacity models, or something like that},
many current forwards- and backwards-in-time simulation software
make these same assumptions, so that Demes is relevant to a range of
widely-used simulators.
\jkcomment{I guess these are also the basic assumptions also shared by
inference methods, which is important. Mabye another sentence here making this
explicit, and listing a few different types of inference methods and giving
citations? This is critical - we want to establish that Demes is the
"middleware" between the inference methods and the simulators, and captures the
basic assumptions made by them all.}
In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
and provide more details on the population genetics models supported by Demes.

@@

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', understood to be a collection of individuals that can be
conveniently modelled using a defined set of rules and parameters
\citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable.
A population is defined as some collection of individuals that exists for
some period of time, and has a well-defined size (i.e., number of individuals)
during that time period. Individuals can move between populations
either according to their ancestor-descendant relationships
or through processes of continuous or pulse migrations.
Few other properties of the populations are specified in the model:
we are concerned primarily with defining the populations, their sizes, and the
movement of individuals between those populations.
\aprcomment{previous sentence changed to allow for selfing/cloning}

@@

Within a population, we implicitly assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously) and the expected frequency (at birth)
of the $i^{th}$ allele currently at frequency $p_i$
one generation in the next generation is $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \cite[][chapter 3]{wakeley2008-hd}.
Further, this model assumes "soft" selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are \textit{independent} of the details of individual fitnesses.
% In the \texttt{Discussion}, we discuss alternative models compatible
% with this framework and outline the expectations for tools implementing
% them.
\aprcomment{As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as defining ``carrying capacities''.}
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][\aprcomment{and more\dots?}]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,jouganous2017inferring,haller2019slim,thornton2019-nu}.
Each has their own unique method of specifying the details of selection and demography.
The goal of \Demes\ is to provide a common method for the latter task.

@@

By allowing the definition of selfing and cloning probabilities, we allow many 
standard models to be defined.
However, by parameterising selfing and cloning as we have, we assume that these
properties of populations can be specified independently from the genetics.
In other words, mutations that cause selfing probabilities to fluctuate within
an epoch are not considered.
\aprcomment{
    Do we still want to say something along these lines?
    ``However, consuming software is free to ignore these parameters
    and model the genetics within
    the framework of a soft selection model.
    We don't view these as limitations. Rather, this is what most people
    are doing most of the time.
    Just now we can finally try to do this part w/fewer errors.''
}
More details of the mathematical properties of selfing and cloning rates
in a coalescent context can be found in \citet{hartfield2016facsexcoal}.

>>>>

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships (i.e., the topology) or by continuous or discrete migration events.
The deomgraphic model itself omits information about genome biology, such as
genome size and architecture, ploidy, and mutation and recombination rates.
It also omits any information about selection operating within populations.
Within populations, allele-frequency dynamics are assumed to
follow the Wright-Fisher model, so that generations are discrete and
non-overlapping and population sizes are independent of mean fitness
(i.e., ``soft'' selection~\citep{christiansen1975hard}).
While these restrictions preclude some population-genetic scenarios of
interest, such as continuous spatial structure
\aprcomment{citations here, and revisit} or
\aprcomment{stochastic carrying-capacity models, or something like that},
many current forwards- and backwards-in-time simulation software
make these same assumptions, so that Demes is relevant to a range of
widely-used simulators.
\jkcomment{I guess these are also the basic assumptions also shared by
inference methods, which is important. Mabye another sentence here making this
explicit, and listing a few different types of inference methods and giving
citations? This is critical - we want to establish that Demes is the
``middleware'' between the inference methods and the simulators, and captures the
basic assumptions made by them all.}
In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
and provide more details on the population genetics models supported by Demes.

@@

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', understood to be a collection of individuals that can be
conveniently modelled using a defined set of rules and parameters
\citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable.
A population is defined as some collection of individuals that exists for
some period of time, and has a well-defined size (i.e., number of individuals)
during that time period. Individuals can move between populations
either according to their ancestor-descendant relationships
or through processes involving migrations.
Few other properties of the populations are specified in the model:
we are concerned primarily with defining the populations, their sizes, and the
movement of individuals between those populations.
\aprcomment{previous sentence changed to allow for selfing/cloning}

@@

Within a population, we implicitly assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously) and the expected frequency (at birth)
of the $i^{th}$ allele currently at frequency $p_i$
one generation in the next generation is $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \cite[][chapter 3]{wakeley2008-hd}.
Further, this model assumes ``soft'' selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are \textit{independent} of the details of individual fitnesses.
% In the \texttt{Discussion}, we discuss alternative models compatible
% with this framework and outline the expectations for tools implementing
% them.
\aprcomment{As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as defining ``carrying capacities''.}
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][\aprcomment{and more\dots?}]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,jouganous2017inferring,haller2019slim,thornton2019-nu}.
Each has their own unique method of specifying the details of selection and demography.
The goal of \Demes\ is to provide a common method for the latter task.

@@

By allowing the definition of selfing and cloning probabilities, we allow many 
standard models to be defined.
However, by parameterising selfing and cloning as we have, we assume that these
properties of populations can be specified independently from the genetics.
In other words, mutations that cause selfing probabilities to fluctuate within
an epoch are not considered.
\aprcomment{
    Do we still want to say something along these lines?
    ``However, consuming software is free to ignore these parameters
    and model the genetics within
    the framework of a soft selection model.
    We don't view these as limitations. Rather, this is what most people
    are doing most of the time.
    Just now we can finally try to do this part w/fewer errors.''
}
\krtcomment{
    I think we want a section in the Discussion outlining some obvious things
    that people can do w/demes that are sitting "just outside" the strict
    specification: hard selection, etc..
}
More details of the mathematical properties of selfing and cloning rates
in a coalescent context can be found in \citet{hartfield2016facsexcoal}.
